---
title: "Regional Income Growth and House Price Dynamics in the Netherlands"
author: 'Vincent, Joris (2890473), Vincent, Thies, Kian & Julius (2898227) '
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Set-up your environment
hoi
```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Regional Income Growth and House Price Dynamics in the Netherlands

Vincent, Joris, Vincent, Thies, Kian & Julius

Tutorial group 3.3 

C. Schouwenaar

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
dataset <- midwest
```
dataset <- midwest

hoge_veerkracht_2024 <- #read.csv("C:/Users/vince/Downloads/hoge-veerkracht-2024.csv", header=FALSE, sep=",")

write.csv(hoge_veerkracht_2024, "data/veerkracht2024.csv")

Inkomen_van_huishoudens__huishoudenskenmerken__regio__indeling_2024__02062026_164030 <- #read.csv("C:/Users/vince/Downloads/Inkomen_van_huishoudens__huishoudenskenmerken__regio__indeling_2024__02062026_164030.csv", header=FALSE, sep=",")

write.csv(Inkomen_van_huishoudens__huishoudenskenmerken__regio__indeling_2024__02062026_164030, "data/inkomenperregio.csv")

<<<<<<< HEAD
Bestaande_koopwoningen <- #read.csv("Bestaande_koopwoningen_gemiddelde_verkoopprijzen_regio_02062026_164630", header = FALSE, sep=",")

write.csv(bestaande_koopwoningen_gemiddelde_verkoopprijzen_regio_02062026_164630, "data/bestaande koopwoningen.csv")
=======
 <<<<<<< HEAD
=======

```{r}

#Bestaande_koopwoningen <- #read.csv("Bestaande_koopwoningen_gemiddelde_verkoopprijzen_regio_02062026_164630", header = FALSE, sep=",")

#write.csv(bestaande_koopwoningen_gemiddelde_verkoopprijzen_regio_02062026_164630, "data/bestaande koopwoningen.csv")

#data_2015 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2015.csv", header=FALSE, sep=";")

write.csv(X2015, "data/data2015.csv")

#data_2016 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2016.csv", header=FALSE, sep=";")

write.csv(X2016, "data/data2016.csv")

#data_2017 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2017.csv", header=FALSE, sep=";")

write.csv(X2017, "data/data2017.csv")

#data_2018 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2018.csv", header=FALSE, sep=";")

write.csv(X2018, "data/data2018.csv")

#data_2019 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2019.csv", header=FALSE, sep=";")

write.csv(X2019, "data/data2019.csv")

#data_2020 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2020.csv", header=FALSE, sep=";")

write.csv(X2020, "data/data2020.csv")

#data_2021 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2021.csv", header=FALSE, sep=";")

write.csv(X2021, "data/data2021.csv")

#data_2022 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2022.csv", header=FALSE, sep=";")

write.csv(X2022, "data/data2022.csv")

#data_2023 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2023.csv", header=FALSE, sep=";")

write.csv(X2023, "data/data2023.csv")

```

```{r}
#prijswoningen <- read.csv(C:\Uni\Year 1\Period 6\Programming for Economics/gem_prijs_woning_prov)

write.csv(Gem_prijs_woning_prov, "goede data/woningprijsprov.csv")
```


## joining

```{r}
data2019 <- read.csv("data/data2019.csv")
data2019$jaar <- 2019
data2020 <- read.csv("data/data2020.csv")
data2020$jaar <- 2020
data2021 <- read.csv("data/data2021.csv")
data2021$jaar <- 2021
data2022 <- read.csv("data/data2022.csv")
data2022$jaar <- 2022
data2023 <- read.csv("data/data2023.csv")
data2023$jaar <- 2023


data2015 <- read.csv("data/data2015.csv")
data2015$jaar <- 2015
data2016 <- read.csv("data/data2016.csv")
data2016$jaar <- 2016

data1516 <- rbind(data2015, data2016)

data2017 <- read.csv("data/data2017.csv")
data2017$jaar <- 2017
data2018 <- read.csv("data/data2018.csv")
data2018$jaar <- 2018

data2019 <- read.csv("data/data2019.csv")
data2019$jaar <- 2019
data2020 <- read.csv("data/data2020.csv")
data2020$jaar <- 2020
data2021 <- read.csv("data/data2021.csv")
data2021$jaar <- 2021
data2022 <- read.csv("data/data2022.csv")
data2022$jaar <- 2022
data2023 <- read.csv("data/data2023.csv")
data2023$jaar <- 2023

data1718 <- rbind(data2017, data2018)

Data15161718 <- rbind(data1516, data1718)

write.csv(Data15161718, "data/data15161718.csv")

datacombined15tot23 <- rbind(Data15161718, data2019, data2020, data2021, data2022, data2023)

write.csv(datacombined15tot23, "data/datacombined15tot23.csv")

datacombined15tot23 <- datacombined15tot23[datacombined15tot23$X != 13, ]

colnames(datacombined15tot23)

nieuwe_rij <- data.frame(X = 13, Regio.s = "Bron: CBS", x.1.000 = NA, X. = NA, X1.000.euro...4 = NA, X1.000.euro...5 = NA, X1.000.euro...6 = NA, X1.000.euro...7 = NA, jaar = NA)
datacombined15tot23 <- rbind(datacombined15tot23, nieuwe_rij)

saveRDS(datacombined15tot23, "goede data/datacombi1523.rds")

```


>>>>>>> aa58531b97baaa294028804c92f44f413d5b2bce
>>>>>>> bac2c3c337e512b99b18e5db5831cf373552a9ab

gemiddelde_inkomens <- #read.csv("inkomen_gemiddeld_", header = FALSE, sep",")

write.csv(inkomen_gemiddeld_, "data/inkomen_gemiddeld_)
<<<<<<< HEAD
 midwest is an example dataset included in the tidyverse package
=======

head(table_86004NED.csv)
nrow(table_86004NED.csv)
nrow(table_86004NED.csv)

table_86004NED.csv <- table_86004NED.csv[1:(nrow(table_86004NED.csv)-4), ]

midwest is an example dataset included in the tidyverse package
>>>>>>> bdbc7d74703b84c02c3eae73a71647a33714c1d8

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning
Data cleaning codering datacombined15tot23goed.
```{r}
names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...6"] <- "gem_st_inkomen_x_100"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...7"] <- "med_st_inkomen_x_100"

<<<<<<< HEAD
names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...8"] <- "gem_bstb_inkomen_x_100"
  
names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...9"] <- "med_bstb_inkomen_x_100"
=======
names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...8"] <- "gem_bstb_inkomen_x_100"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...9"] <- "med_bstb_inkomen_x_100"
>>>>>>> c6518e6db0c39fa1f561bcda18b5c97d5399a3c5

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...9"] <- "med_bstb_inkomen_x_1000"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "Regio.s"] <- "Provincies"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "`nr_huishoudens"] <- "ant_hh_p_provincie"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X"] <- "Provincie_nr"


saveRDS(datacombined15tot23goed, "goede data/datacombined15tot23goed.rds")

datacombined15tot23goed$X. <- NULL
datacombined15tot23goed$...1 <- NULL

```
Data cleaning codering woningprijsprov
```{r}
woningprijsprov$`1995` <- format(woningprijsprov$`1995`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2000` <- format(woningprijsprov$`2000`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2005` <- format(woningprijsprov$`2005`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2010` <- format(woningprijsprov$`2010`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2015` <- format(woningprijsprov$`2015`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2020` <- format(woningprijsprov$`2020`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2021` <- format(woningprijsprov$`2021`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2022` <- format(woningprijsprov$`2022`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2023` <- format(woningprijsprov$`2023`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2024` <- format(woningprijsprov$`2024`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2025` <- format(woningprijsprov$`2025`, big.mark = ".", scientific = FALSE)
saveRDS(woningprijsprov, "goede data/gem_huizenprijs_per_provincie.RDS")

```


Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
