---
title: "Regional Income Growth and House Price Dynamics in the Netherlands"
author: 'Vincent, Joris (2890473), Vincent, Thies, Kian & Julius (2898227) '
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Set-up your environment
```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
```

```{r packages, include=FALSE}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Regional Income Growth and House Price Dynamics in the Netherlands

Vincent, Joris, Vincent, Thies, Kian & Julius

Tutorial group 3.3 


# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

The Dutch housing market has been under pressure for several years now. Between 2013 and 2023 the average price of houses has more than doubled, while the income growth in the same period considerably lagged behind. The increasing gap between housing costs and purchasing power is not only an economic phenomenon, but also a social problem; it determines who does have access to a stable living situation and who doesn’t.

It’s striking that this development differs significantly across regions in the Netherlands. In the Randstad, prices increased much more than in the peripheral regions like Groningen or Zeeland, while the income development per region differs significantly too. Factors such as access to big employers, education level and local economic structure play an important role in this shift. And it is because of this regional variety that it’s interesting to investigate the relation between income and house prices. 

This project investigates the relationship between regional income growth and house price dynamics in Dutch provinces, based on CBS-data and the programming tool R. The goal is to discover in which regions the housing market detached most from the local income development.


# Part 2 - Data Sourcing

## 2.1 Load in the data
we used 9 different data sets and combined we did this using this code for each
first we cleaned the already existing datasets. aand combined everything together using this code: datacombined15tot23 <- rbind(Data15161718, data2019, data2020, data2021, data2022, data2023).

```{r include=FALSE}

#Bestaande_koopwoningen <- #read.csv("Bestaande_koopwoningen_gemiddelde_verkoopprijzen_regio_02062026_164630", header = FALSE, sep=",")

#write.csv(bestaande_koopwoningen_gemiddelde_verkoopprijzen_regio_02062026_164630, "data/bestaande koopwoningen.csv")

#data_2015 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2015.csv", header=FALSE, sep=";")
X2015 <- read.csv("data/data2015.csv")
write.csv(X2015, "data/data2015.csv", row.names = FALSE)

#data_2016 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2016.csv", header=FALSE, sep=";")
X2016 <- read.csv("data/data2016.csv")
write.csv(X2016, "data/data2016.csv", row.names = FALSE)

#data_2017 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2017.csv", header=FALSE, sep=";")
X2017 <- read.csv("data/data2017.csv")
write.csv(X2017, "data/data2017.csv", row.names = FALSE)

#data_2018 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2018.csv", header=FALSE, sep=";")
X2018 <- read.csv("data/data2018.csv")
write.csv(X2018, "data/data2018.csv", row.names = FALSE)

#data_2019 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2019.csv", header=FALSE, sep=";")
X2019 <- read.csv("data/data2019.csv")
write.csv(X2019, "data/data2019.csv", row.names = FALSE)

#data_2020 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2020.csv", header=FALSE, sep=";")
 X2020 <- read.csv("data/data2020.csv")
write.csv(X2020, "data/data2020.csv", row.names = FALSE)

#data_2021 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2021.csv", header=FALSE, sep=";")
X2021 <- read.csv("data/data2021.csv")
write.csv(X2021, "data/data2021.csv", row.names = FALSE)

#data_2022 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2022.csv", header=FALSE, sep=";")
X2022 <- read.csv("data/data2022.csv")
write.csv(X2022, "data/data2022.csv", row.names = FALSE)

#data_2023 <- read.csv("C:/Uni/Year 1/Period 6/Programming for Economics/2023.csv", header=FALSE, sep=";")
X2023 <- read.csv("data/data2023.csv")
write.csv(X2023, "data/data2023.csv", row.names = FALSE)

```

```{r include=FALSE}
# prijswoningen <- read.csv(C:\Uni\Year 1\Period 6\Programming for Economics/gem_prijs_woning_prov)
# Gem_prijs_woning_prov <- read.csv("goede data/woningprijsprov.csv")
# write.csv(Gem_prijs_woning_prov, "goede data/woningprijsprov.csv")
```

```{r}
Bev_dichtheid <- read.csv("C:/Users/vince/Downloads/Regionale_kerncijfers_Nederland_10062026_162325.csv", header=FALSE, sep=";")
write.csv(Bev_dichtheid, "goede data/bevolkingsdichtheid.csv")
```


## joining

```{r include=FALSE}


data2015 <- read.csv("data/data2015.csv")
data2015$jaar <- 2015
data2016 <- read.csv("data/data2016.csv")
data2016$jaar <- 2016

data2015$X.5 <- NULL
data2016$X.5 <- NULL

data2015$X.7 <- NULL
data2016$X.7 <- NULL

data1516 <- rbind(data2015, data2016)

data2017 <- read.csv("data/data2017.csv")
data2017$jaar <- 2017
data2018 <- read.csv("data/data2018.csv")
data2018$jaar <- 2018

data2019 <- read.csv("data/data2019.csv")
data2019$jaar <- 2019

data2020 <- read.csv("data/data2020.csv")
data2020$jaar <- 2020
data2021 <- read.csv("data/data2021.csv")
data2021$jaar <- 2021
data2022 <- read.csv("data/data2022.csv")
data2022$jaar <- 2022
data2023 <- read.csv("data/data2023.csv")
data2023$jaar <- 2023

data2017$X.5 <- NULL
data2018$X.5 <- NULL
data2019$X.5 <- NULL
data2020$X.5 <- NULL
data2021$X.5 <- NULL
data2022$X.5 <- NULL
data2023$X.5 <- NULL

data1718 <- rbind(data2017, data2018)

Data15161718 <- rbind(data1516, data1718)

write.csv(Data15161718, "data/data15161718.csv")

datacombined15tot23 <- rbind(Data15161718, data2019, data2020, data2021, data2022, data2023)

write.csv(datacombined15tot23, "data/datacombined15tot23.csv")

datacombined15tot23 <- datacombined15tot23[datacombined15tot23$X != 13, ]

colnames(datacombined15tot23)

saveRDS(datacombined15tot23, "goede data/datacombi1523.rds")

```

## 2.2 Provide a short summary of the dataset(s)

For this study two datasets from the Centraal Bureau voor de Statistiek (CBS) were used. The first dataset, Bestaande koopwoningen; verkoopprijzen prijsindex 2020=100 (85773NED), contains data on the variables ‘gemiddelde verkoopprijs van bestaande koopwoningen en bijbehorende prijsindex op provinciaal niveau’. The second dataset, inkomen van huishoudens; huishoudenskenmerken, regio (86004NED), contains information on the ‘gemiddelde inkomen van huishoudens, op provinciaal niveau’.

Both datasets are available on CBS StatLine. The dataset on the prices of houses spans from 2015 until 2025, the dataset on income only covers 2015 until 2023 (In order to see other years or regions, the filter option should be used to select the desired years).  For the analysis in this research the overlapping years of 2015-2023 are studied. 


## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?


# Part 3 - Quantifying

## 3.1 Data cleaning
Data cleaning datacombined15tot23goed.
we mainly used this code to clean this data set:
names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...6"] <- "gem_st_inkomen_x_100"
ETC.

```{r include=FALSE}
datacombined15tot23goed <- read.csv("goede data/datacombined15tot23goed.csv")
names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...6"] <- "gem_st_inkomen_x_100"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...7"] <- "med_st_inkomen_x_100"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...8"] <- "gem_bstb_inkomen_x_100"
  
names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...9"] <- "med_bstb_inkomen_x_100"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...8"] <- "gem_bstb_inkomen_x_100"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...9"] <- "med_bstb_inkomen_x_100"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X1.000.euro...9"] <- "med_bstb_inkomen_x_1000"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "Regio.s"] <- "Provincies"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "`nr_huishoudens"] <- "ant_hh_p_provincie"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "X"] <- "Provincie_nr"

names(datacombined15tot23goed)[names(datacombined15tot23goed) == "x.1.000"] <- "Nr_huishoudens"

write.csv(datacombined15tot23goed, "goede data/datacombined15tot23goed.csv", row.names = FALSE)

saveRDS(datacombined15tot23goed, "goede data/datacombined15tot23goed.rds")

datacombined15tot23goed <- datacombined15tot23goed[datacombined15tot23goed$Provincies != "Bron: CBS", ]

datacombined15tot23goed$X. <- NULL
datacombined15tot23goed$...1 <- NULL

```
Data cleaning woningprijsprov:
we mainly used this code to clean this data set:
woningprijsprov$`1995` <- format(woningprijsprov$`1995`, big.mark = ".", scientific = FALSE)
```{r include=FALSE}
woningprijsprov <- read.csv("goede data/goedehuizenprijsenperprov.csv")
woningprijsprov$`1995` <- format(woningprijsprov$`1995`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2000` <- format(woningprijsprov$`2000`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2005` <- format(woningprijsprov$`2005`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2010` <- format(woningprijsprov$`2010`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2015` <- format(woningprijsprov$`2015`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2020` <- format(woningprijsprov$`2020`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2021` <- format(woningprijsprov$`2021`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2022` <- format(woningprijsprov$`2022`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2023` <- format(woningprijsprov$`2023`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2024` <- format(woningprijsprov$`2024`, big.mark = ".", scientific = FALSE)
woningprijsprov$`2025` <- format(woningprijsprov$`2025`, big.mark = ".", scientific = FALSE)
woningprijsprov$...1 <- NULL

saveRDS(woningprijsprov, "goede data/gem_huizenprijs_per_provincie.RDS")

```

```{r include=FALSE}
gem_huizenprijs_per_provincie <- readRDS("goede data/gem_huizenprijs_per_provincie.RDS")
write.csv(gem_huizenprijs_per_provincie, "goede data/goedehuizenprijsenperprov.csv", row.names = FALSE)
```


```{r}
library(tidyverse)

# Data inladen
bevolking <- read.csv(
  "data/Regionale_kerncijfers_Nederland_10062026_162325.csv",
  sep = ";",
  header = TRUE,
  strip.white = TRUE,
  fileEncoding = "UTF-8-BOM"
)

# Kolom hernoemen
names(bevolking)[3] <- "bevolkingsdichtheid_km2"

# Opschonen en breed maken
bevolking_breed <- bevolking %>%
  filter(Regio.s != "Nederland") %>%
  mutate(
    Regio.s = gsub(" \\(PV\\)", "", Regio.s),
    bevolkingsdichtheid_km2 = as.numeric(bevolkingsdichtheid_km2)
  ) %>%
  pivot_wider(
    names_from = Perioden,      # jaartallen worden kolomnamen
    values_from = bevolkingsdichtheid_km2
  ) %>%
  arrange(Regio.s)              # provincies alfabetisch sorteren

# Bekijken
View(bevolking_breed)
```



Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1
creating new variable 1, we have created the average disposable income of each province over 8 years. using the data set: gem_inkomen_provincie, We used the following code for this:
gem_inkomen_provincie <- aggregate(gem_bstb_inkomen_x_100 ~ Provincies, 
                                    data = datacombined15tot23goed, 
                                    FUN = mean)

```{r gen_var1, include=FALSE}

gem_inkomen_provincie <- aggregate(gem_bstb_inkomen_x_100 ~ Provincies, 
                                    data = datacombined15tot23goed, 
                                    FUN = mean)

write.csv(gem_inkomen_provincie, "goede data/gem_inkomen_provincie.csv", row.names = FALSE)

colnames(gem_inkomen_provincie)[colnames(gem_inkomen_provincie) == "gem_bstb_inkomen_x_100"] <- "gem_bstb_inkomen_per_jaar_x_100"

write.csv(gem_inkomen_provincie, "goede data/gem_inkomen_provincie.csv", row.names = FALSE)
```

Variable 2
creating new variable 2 using the original dataset: datacombined15tot23goed, in which we represented the weighted averages of income for different provinces: inside and outside the Randstad.

adjust the format of the current cleaned dataset in order to continue working with a renewed dataset, and also a renewed dataset, because for this new variable we actually need to merge our two datasets.
```{r gen_var2}
woningprijsprov <- read.csv("goede data/goedehuizenprijsenperprov.csv", check.names = FALSE)

woningprijsprov <- woningprijsprov[!grepl("Bron", woningprijsprov$`Regio.s`), ]

woningprijsprov$Onderwerp <- NULL
woningprijsprov$`...4` <- NULL

woningprijsprov_lang <- reshape(woningprijsprov,
                                varying = c("2015", "2020", "2021", "2022", "2023"),
                                v.names = "gem_huizenprijs",
                                timevar = "jaar",
                                times = c(2015, 2020, 2021, 2022, 2023),
                                direction = "long")

woningprijsprov_lang$id <- NULL

write.csv(woningprijsprov_lang, "goede data/goedehuizenprijsenperprov_lang.csv", row.names = FALSE)
```

## 3.3 Visualize temporal variation

```{r echo=FALSE}
library(tidyverse)
library(scales)

# Data inladen
df <- read.csv("goede data/randstad_vs_buiten_randstad.csv", stringsAsFactors = FALSE)

# Cumulatieve groei berekenen
df_plot <- df %>%
  arrange(regio, jaar) %>%
  group_by(regio) %>%
  mutate(
    groei_huizen = (gem_huizenprijs / first(gem_huizenprijs)) - 1,
    groei_inkomen = (gem_bstb_inkomen_euro / first(gem_bstb_inkomen_euro)) - 1
  ) %>%
  ungroup()

# Data omzetten naar lang formaat
df_long <- df_plot %>%
  pivot_longer(cols = c(groei_huizen, groei_inkomen),
               names_to = "type",
               values_to = "groei") %>%
  mutate(type = recode(type,
                       "groei_huizen" = "Huizenprijs",
                       "groei_inkomen" = "Inkomen"))

# Functie voor grafiek
maak_grafiek <- function(data, titel) {
  ggplot(data, aes(x = jaar, y = groei, color = type, linetype = type)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_x_continuous(breaks = c(2020, 2021, 2022, 2023)) +
    scale_color_manual(values = c("Huizenprijs" = "steelblue", "Inkomen" = "orange")) +
    scale_linetype_manual(values = c("Huizenprijs" = "solid", "Inkomen" = "dashed")) +
    labs(title = titel, x = "Jaar", y = "Groei t.o.v. 2020",
         color = "", linetype = "") +
    theme_minimal() +
    theme(legend.position = "top")
}

# Automatisch beide grafieken tonen
grafieken_tonen <- function() {
  print(maak_grafiek(
    df_long %>% filter(regio == "Randstad (incl. Flevoland)"),
    "Groei Randstad t.o.v. 2020"
  ))
  print(maak_grafiek(
    df_long %>% filter(regio == "Buiten Randstad"),
    "Groei Buiten Randstad t.o.v. 2020"
  ))
}

grafieken_tonen()
```

## 3.4 Visualize spatial variation

```{r visualise_map, echo=FALSE}
install.packages(c("sf", "ggplot2"))
library(sf)
library(ggplot2)
nl_map <- st_read("https://cartomap.github.io/nl/wgs84/provincie_2024.geojson")
nl_map$statnaam
my_data <- data.frame(
  Provincies = c("Groningen", "Fryslân", "Limburg", "Drenthe", "Overijssel",
                 "Zeeland", "Zuid-Holland", "Gelderland", "Flevoland",
                 "Noord-Brabant", "Noord-Holland", "Utrecht"),
  mean_income = c(375.5556, 418.3333, 423.7778, 440.7778, 443.6667,
                  445.1111, 453.8889, 454.7778, 461.2222, 466.7778,
                  474.8889, 493.5556)
)
nl_map <- merge(nl_map, my_data, by.x = "statnaam", by.y = "Provincies")
ggplot(nl_map) +
  geom_sf(aes(fill = mean_income)) +
  scale_fill_distiller(
    palette = "Blues",
    direction = 1,        # 1 = darker means higher value
    name = "Mean spendable income (x100 €)"
  
  ) +
  theme_void() 
  labs(
    title = "Mean Spendable Income per Province",
    subtitle = "Average over 2015-2023"
  )
  
  
  library(sf)
library(ggplot2)

# Stap 1: Kaart laden
nl_map <- st_read("https://cartomap.github.io/nl/wgs84/provincie_2024.geojson")

# Stap 2: Jouw data
my_data2 <- data.frame(
  Provincies = c("Groningen", "Fryslân", "Drenthe", "Overijssel", "Flevoland",
                 "Gelderland", "Utrecht", "Noord-Holland", "Zuid-Holland",
                 "Zeeland", "Noord-Brabant", "Limburg"),
  huizenprijs = c(265840.0, 274763.0, 296620.4, 307718.0, 329630.8,
                  353163.8, 423367.6, 446549.2, 355059.0,
                  280299.6, 360750.2, 283216.8)
)

# Stap 3: Samenvoegen
nl_map <- merge(nl_map, my_data2, by.x = "statnaam", by.y = "Provincies")

# Stap 4: Kaart plotten
ggplot(nl_map) +
  geom_sf(aes(fill = huizenprijs)) +
  scale_fill_distiller(
    palette = "Greens",
    direction = 1,
    name = "Gem. huizenprijs (€)",
    labels = scales::comma
  ) +
  theme_void() 
  labs(
    title = "Gemiddelde Huizenprijs per Provincie",
    subtitle = "Gemiddelde over de laatste jaren"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
